Parsing Postfix Log Files

### 2009-04-26 doc/logparser-architecture.tex
Changes based on proofreading chapters 4 and 5.

To avoid cluttering the description of the parser architecture with the
details involved in implementing a parser for Postfix log files, each topic
has been given its own chapter.  This chapter presents the architecture
developed for this project, beginning with the overall architecture and
design, followed by detailed documentation of the three components of the
architecture: Framework, Actions, and Rules.  This chapter centers on the
theoretical, implementation-independent aspects of the architecture; the
practical difficulties of writing a parser for Postfix log files are
covered in detail in \sectionref{Postfix Parser Implementation}.

@@

It should be clear from the earlier Postfix background (\sectionref{postfix
background}) that log files produced by Postfix may vary widely from host
to host.  With this in mind, one of the architecture's design aims was to
make parsing new inputs as effortless as possible, to enable administrators
to properly parse their own log files.  The solution developed is to divide
the architecture into three parts: Framework, Actions, and Rules.  Each
will be discussed separately, but first an overview:

@@

    \item [Actions] Each action performs the processing required for a
        single \textit{category\/} of inputs, e.g.\ rejecting a delivery
        attempt.  Actions are invoked to process an input once it has been
        recognised by a rule.

    \item [Rules]  Rules are responsible for classifying inputs: each rule
        recognises one input \textit{variant\/} (a single input category
        may have many input variants).  Each rule also specifies the action
        to be invoked when an input has been recognised; rules thus provide
        an extensible method of associating inputs with actions.

@@

For each input, the framework tries each rule in turn until it finds a rule
that recognises the input, then invokes the action specified by that rule.
If the input is not recognised by any of the rules, the framework issues a
warning; the framework will usually continue parsing after this, although
for some inputs stopping immediately may be preferable.

@@

\begin{table}[thbp]
    \caption{Similarities between ATN and this architecture}
    \empty{}\label{Similarities between ATN and this architecture}
    \centering{}
    \begin{tabular}[]{lll}
        \tabletopline{}%
        \acronym{ATN}   & Architecture  & Similarity                  \\
        \tablemiddleline{}%
        Networks        & Framework     & Determines the sequence
                                          of transitions              \\
                        &               & or actions that
                                          constitutes a valid input.  \\
        Transitions     & Actions       & Assemble data and
                                          impose conditions the       \\
                        &               & input must meet to be
                                          accepted as valid.          \\
        Abbreviations   & Rules         & Responsible for
                                          classifying input.          \\
        \tablebottomline{}%
    \end{tabular}
\end{table}

@@

The framework manages the parsing process and provides support functions
for the actions, freeing the programmers writing actions to concentrate on
writing productive code.  It links actions and rules, allowing either to be
improved independently of the other, and allows new rules to be written
without needing changes to the source code of a parser.  The framework is
the core of the architecture and is deliberately quite simple: the rules
deal with the variation in inputs, and the actions deal with the
intricacies and complications encountered during parsing.  The function
that finds the rule that recognises the input and invokes the requested
action can be expressed in pseudo-code as shown in \figureref{Pseudo-code
describing how inputs are recognised}.

\begin{figure}[thbp]
    \caption{Pseudo-code describing how inputs are recognised}
    \empty{}\label{Pseudo-code describing how inputs are recognised}
\begin{verbatim}
INPUT:
for each input {
    for each rule defined by the user {
        if this rule recognises the input {
            invoke the action specified by the rule
            next INPUT
        }
    }
    warn the user that the input was not recognised
}
\end{verbatim}
\end{figure}

@@

    \item [Register actions]  Each action needs to be registered with the
        framework so that the framework knows about it; the list of
        registered actions will also be used when validating rules.

    \item [Load and validate rules]  The framework loads the rules from
        wherever they are stored: a simple file, a database, or possibly
        even a web server or other network service --- though that would
        have serious security implications.  It validates each rule to
        catch problems as early in the parsing process as possible; the
        checks will be implementation-specific to some extent, but will
        generally include some of the following:

@@

        Some optimisation steps may also be performed during loading of
        rules, as discussed in \sectionref{parser efficiency}.

    \item [Convert physical inputs to logical inputs] Each rule recognises
        a single input at a time: there is no facility for rules to consume
        more input or push unused input back onto the input stream,
        although actions may use cascaded parsing (explained in
        \sectionref{actions in architecture}) to effectively push input
        back onto the input stream.  A physical input (e.g.\ a single line
        read from the input stream) may contain multiple or partial logical
        inputs, and the framework must pre-process these to provide a
        logical input for the rules to recognise.  This simplifies the
        rules and actions considerably, at the cost of added complexity in
        the framework; during the design phase it was decided that it was
        easier to deal with the problem of parsing multiple or partial
        inputs once, rather than dealing with it in every rule and action.
        This is trivial for Postfix log files because they have a
        one-to-one mapping between physical and logical inputs; mapping
        between physical and logical inputs may be more difficult for other
        types of input.  Some input types may require pre-processing
        equivalent to parsing the physical inputs; in those cases the
        framework present the entire input stream at once, and discard the
        recognised portion.

    \item [Classify the input]  The pseudo-code in \figureref{Pseudo-code
        describing how inputs are recognised} shows how rules are
        successively tried until one is found that recognises the current
        input.  That pseudo-code is very simple: there may be efficiency
        concerns (\sectionref{parser efficiency}), rule conditions
        (\sectionref{attaching conditions to rules}), or rule priorities
        (\sectionref{rules in architecture}) that complicate the process.

    \item [Invoke actions]  Once a rule has been found that recognises the
        current input, the specified action must be invoked.  The framework
        marshals the data extracted by the rule, invokes the action, and
        repeats the parsing process if cascaded parsing (see
        \sectionref{actions in architecture}) is used.

    \item [Shared storage]  Parsers commonly need to save some state
        information about the input being parsed, e.g.\ a compiler tracking
        which variables are in lexical scope as it moves from one lexical
        block to another.  The framework provides shared storage to deal
        with this and any other storage needs the actions have.  Actions
        may need to exchange data to correctly parse the input, e.g.\
        setting or clearing flags, maintaining a list of previously used
        identifiers, or ensuring at a higher level that the input being
        parsed meets requirements.

    \item [Save and load state]  The architecture can save the contents of
        the shared storage it provides for actions, and reload it later, so
        that information is not lost between parsing runs.  \parsername{}
        uses this functionality because mails may take some time to deliver
        and thus have their log lines split between log files; a compiler
        might store data structures it builds as it parses different files.

    \item [Specialised support functions]  Actions may need support or
        utility functions; the framework may be a good location for support
        functions, but if another way exists to make those functions
        available to all actions it may be preferable to use it instead,
        maintaining a clear separation of concerns.

@@

Each action is a separate procedure written to process a particular
category of input, e.g.\ rejection of a delivery attempt.  There may be
many input variants within one input category; in general each action will
handle one input category, with each rule recognising one input variant.
It is anticipated that parsers based on this architecture will have many
actions, and each action may be specified by many rules, with the aim of
having simple rules and tightly focused actions.  An action may need to
process different input variants in slightly different ways, but large
changes in processing within an action indicate the need for a new action
and a new category of input; if an action becomes overly complicated it
starts to turn into a monolithic parser, with too much logic contained in a
single procedure.

The ability to easily add special purpose actions to deal with difficulties
and new requirements that are discovered during parser development is one
of the strengths of this architecture.  When a new requirement arises an
independent action can be written to satisfy it, instead of writing a
single monolithic function that processes every input and must be modified
to support any change in behaviour, with all the attendant risks of
adversely affecting the existing parsing.  Sometimes the new action will
require the cooperation of other actions, e.g.\ to set or check a flag, so
actions are not always self-contained, but there will still be a far lower
degree of coupling and interdependency than in a monolithic parser.

During development of \parsername{} it became apparent that in addition to
the obvious variety in log lines there were many complications to be
overcome.  Some were the result of deficiencies in Postfix's logging, and
some of those deficiencies were rectified by later versions of Postfix,
e.g.\ identifying bounce notifications (\sectionref{identifying bounce
notifications}); others were due to the vagaries of process scheduling,
client behaviour, and administrative actions.  All were successfully
accommodated in \parsername{}: adding new actions was enough to overcome
several of the complications; others required modifications to a single
existing action to work around a difficulty; the remainder were resolved by
adapting existing actions to cooperate and exchange extra data (via the
framework), changing their behaviour as appropriate based on that extra
data.  Every architecture should aim to make the easy things easy and the
hard things possible; the successful implementation of \parsername{}
demonstrates that this architecture achieves that aim.

@@

@@

Rules are responsible for categorising inputs: each rule should recognise
one and only one input variant; an input category with multiple input
variants should have multiple rules, one for each variant.  Rules will
typically use a regex when recognising inputs, but other approaches may
prove useful for some applications, e.g.\ comparing fixed strings to the
input, or checking the length of the input; for the remainder of this
thesis it will be assumed that a regex is used.  Each rule must specify at
least the regex to recognise inputs and the action to invoke when
recognition is successful, but implementations are free to add any other
attributes they require; \sectionref{rules table} describes the attributes
used in \parsername{}, and some generally useful attributes will be
discussed later in this section.

Using the rules is simple: the first rule to recognise the input determines
the action that will be invoked; there is no backtracking to try alternate
rules; no attempt is made to pick a \textit{best\/} rule.
\sectionref{attaching conditions to rules} contains an example which
requires that the rules are used in a specific order to correctly parse the
input, so a mechanism is needed to allow the author of the rules to specify
that ordering.  The framework should sort the rules according to a priority
attribute specified by the rule author, allowing fine-grained control over
the order that rules are used in.  The priority attribute may be
implemented as a number, or as a range of values, e.g.\ low, medium, and
high, or in a different fashion entirely if it suits the implementation.
Rule ordering for efficiency is a separate topic that is covered in
\sectionref{rule ordering for efficiency}; overlapping rules are discussed
in \sectionref{overlapping rules in architecture}.

@@

Decoupling the rules from the actions and framework enables other rule
management approaches to be used, e.g.\ instead of manually editing
existing rules or adding new rules, machine learning techniques could be
used to automate the process.  If this approach was taken the choice of
machine learning technique would be constrained by the size of typical data
sets (see \sectionref{parser efficiency}).  Techniques requiring the full
data set when training would be impractical; Instance Based
Learning~\cite{instance-based-learning} techniques that automatically
determine which inputs from the training set are valuable and which inputs
can be discarded might reduce the data required to a manageable size.  A
parser could also dynamically create new rules in response to certain
inputs, e.g.\ parsing a subroutine declaration could cause a rule to be
created that parses calls to that subroutine, checking that the arguments
used agree with the subroutine's signature.  These avenues of research and
development have not been pursued by the author, but the architecture
allows them to easily be undertaken independently.

@@

Rules can have conditions attached that will be evaluated by the framework
before attempting to use a rule to recognise an input: if the condition is
true the rule will be used, if not the rule will be skipped.  Conditions
can be as simple or complex as the parser requires, though naturally as the
complexity rises so too does the difficulty in understanding how everything
interacts.  The framework has to evaluate each condition, so as the
complexity of conditions increases so will the complexity of the code
required to evaluate them.  Beyond a certain level of complexity,
conditions should probably be written in a proper programming language,
e.g.\ taking advantage of dynamic languages' support for evaluating code at
run-time, or embedding a language like Lua into the parser.  If an
implementation is going to use conditions so complex that they will require
a programming language capable of implementing a Turing machine, the design
may need to be revisited, including the decision to use this architecture
--- there may be other architectures more suitable.

Conditions that examine the input will be the easiest to understand,
because they can be understood in isolation; they do not depend on
variables set by actions or other rules.  Conditions that examine the input
can be complex if required, but simple conditions can be quite useful too,
e.g.\ every Postfix log line contains the name of the Postfix component
that produced it, so every rule used in \parsername{} has a condition
specifying the component whose log lines it recognises, reducing the number
of rules that will be used when classifying a log line (see
\sectionref{rules in implementation} for details) and increasing the chance
that the log line will be correctly recognised.

@@

An example of how rule conditions can be useful is parsing C-style
comments, which start with \texttt{/*} and end with \texttt{*/}; the start
and end tokens can be on one line, or may have many lines between them.
\Tableref{Rules to parse C-style comments} shows the regexes, conditions,
and state changes of the four rules required to parse C-style comments.
These rules are somewhat simplified, e.g.\ rules one and two will recognise
the comment start token even if it is within a quoted string, which is
incorrect; the rules could be extended with new rules to parse quoted
strings, excluding comments from within strings.  Rules one and two will be
used when the parser is parsing code, not comments: rule one recognises a
comment that is contained within one line and leaves the parser's state
unchanged; rule two recognises the start of a comment and changes the
parser's state to parsing comments instead of parsing code.  Rules three
and four will be used when the parser is parsing a comment: rule three
recognises the end of a comment and switches the parser's state back to
parsing code; rule four recognises a comment line without an end token and
keeps the parser's state unchanged.  It is important that the rules are
applied in the order listed in \tableref{Rules to parse C-style comments}
because rule 2 overlaps with rule 1, so rule 1 must be used first to
correctly parse comments that are contained within one line;
\sectionref{rules in architecture} has explained how this is achieved.
\sectionref{overlapping rules in architecture} will discuss the benefits
and difficulties of using overlapping rules.

\begin{table}[thbp]
    \caption{Rules to parse C-style comments}
    \empty{}\label{Rules to parse C-style comments}
    \begin{tabular}{llll}
        \tabletopline{}%
        No.   & Regex             & Condition                 & Variable Changes          \\
        \tablemiddleline{}%
        1     & \verb!/\*.*?\*/!  & state == parsing code     & state = parsing code      \\
        2     & \verb!/\*.*!      & state == parsing code     & state = parsing comment   \\
        3     & \verb!.*\*/!      & state == parsing comment  & state = parsing code      \\
        4     & \verb!.*!         & state == parsing comment  & state = parsing comment   \\
        \tablebottomline{}%
    \end{tabular}

    Note that ``state'' is merely a descriptive name for the variable; the
    variable can called anything at all.

@@

When adding new rules, the rule author must be aware that the new rule may
overlap with one or more existing rules, i.e.\ some inputs could be parsed
by more than one rule.  Unintentionally overlapping rules lead to
inconsistent parsing and data extraction because the first rule to
recognise the input wins, and the order in which rules are used might
change between parser invocations.  Overlapping rules are frequently a
requirement, allowing a more specific rule to recognise some inputs and a
more general rule to recognise the remainder, e.g.\ separating
\acronym{SMTP} delivery to specific sites from \acronym{SMTP} delivery to
the rest of the world.  Using overlapping rules simplifies both the general
rule and the more specific rule; additionally rules from different sources
can be combined with a minimum of prior cooperation or modification
required.  Overlapping rules should have a priority attribute to specify
their relative ordering; negative priorities may be useful for catchall
rules.  The architecture does not try to detect overlapping rules: that
responsibility is left to the author of the rules.

Overlapping rules can be detected by visual inspection, or a program may be
written to analyse the regexes in a ruleset.  Traditional regexes are
equivalent in computational power to \acronym{FA} and can be converted to
\acronym{FA}, so regex overlap can be detected by finding a non-empty
intersection of two \acronym{FA}\@.  Perl 5.10 regexes~\cite{perlre} are
more powerful than traditional regexes: it is possible to match correctly
balanced brackets nested to an arbitrary depth, e.g.\
\verb!/^[^<>]*(<(?:(?>[^<>]+)|(?1))*>)[^<>]*$/!  matches
\verb!z<123<pq<>rs>j<r>ml>s!.  Matching balanced brackets requires the
regex engine to maintain state on a stack, so Perl 5.10 regexes are
equivalent in computational power to \acronym{PDA}; detecting overlap may
require calculating the intersection of two \acronym{PDA} instead of two
\acronym{FA}.  \acronym{PDA} intersection is not closed, i.e.\ the result
cannot always be implemented using a \acronym{PDA}, so intersection may be
intractable sometimes, e.g.:
$a^{*}b^{n}c^{n}~\cap~a^{n}b^{n}c^{*}~\rightarrow~a^{n}b^{n}c^{n}$.
Detecting overlap amongst $n$ regexes requires calculating
$\frac{n\left(n-1\right)}{2}$ intersections, resulting in
$O\left(n^{2}x\right)$ complexity, where $x$ is the cost of calculating
\acronym{FA} or \acronym{PDA} intersection.  This is certainly not a task
to be performed every time a parser runs: naive detection of overlap
amongst \parsernames{} \numberOFrules{} rules would require calculating
\numberOFruleINTERSECTIONS{} intersections.

@@

In the worst case $|o|$ above will be equal to $n^{2}$.  For this approach
to be more efficient than the naive approach, $y$ must be significantly
lower than $x$.  If $y$ is higher than $x$ then the checks for overlap
should be performed in the opposite order: check for regex overlap first,
then check for condition overlap only between pairs of rules with
overlapping regexes.

@@

It is possible to write pathological regexes, which fall into two main
categories: regexes that match inputs they should not, and regexes that
consume excessive amounts of CPU time during matching.  Defining a regex
that matches inputs it should not is trivial: \verb!/^/! matches the start
of every input.  This regex would be found by a tool that detects
overlapping rules, and would easily be noticed by visual inspection, but
more complex regexes would be harder to find.  Regexes that match inputs
more than they should are a problem not because of excessive resource
usage, but because they may prevent the correct rule from recognising the
input.  If an adaptive ordering system is used that prioritises rules that
frequently recognise inputs (see \sectionref{rule ordering for
efficiency}), then a rule with a regex that matches inputs it should not
may be promoted up through the list, displacing an increasing number of
correct rules as it rises.

>>>>

To avoid cluttering the description of the parser architecture with the
details of implementing a parser for Postfix log files, each topic has been
given its own chapter.  This chapter presents the architecture developed
for this project, beginning with the overall architecture and design,
followed by detailed documentation of the three components of the
architecture: Framework, Actions, and Rules.  This chapter is focused on
the theoretical, implementation-independent aspects of the architecture;
the practical difficulties of writing a parser for Postfix log files are
covered in detail in \sectionref{Postfix Parser Implementation}.

@@

It should be clear from the earlier Postfix background (\sectionref{postfix
background}) that Postfix log files may vary widely from host to host.
With this in mind, one of the architecture's design aims was to make
parsing new inputs as effortless as possible, to enable administrators to
properly parse their own log files.  The solution developed is to divide
the architecture into three parts: Framework, Actions, and Rules.  Each
will be documented separately, but first an overview:

@@

    \item [Actions] Each action performs the processing required for a
        single \textit{category\/} of inputs, e.g.\ rejection of a delivery
        attempt.  Actions are invoked to process an input once it has been
        recognised by a rule.

    \item [Rules]  Rules are responsible for classifying inputs: each rule
        recognises one input \textit{variant\/} --- a single input category
        may have many input variants.  Each rule also specifies the action
        to be invoked when an input has been recognised; rules thus provide
        an extensible method of associating inputs with actions.

@@

For each input, the framework tries each rule in turn until it finds a rule
that recognises the input, then invokes the action specified by that rule.
If the input is not recognised by any of the rules, the framework issues a
warning; the framework will usually continue parsing after this, although
some parsers might prefer to stop immediately.

@@

\begin{table}[thbp]
    \caption{Similarities between ATN and this architecture}
    \empty{}\label{Similarities between ATN and this architecture}
    \centering{}
    \begin{tabular}[]{lll}
        \tabletopline{}%
        \acronym{ATN}   & Architecture  & Similarity                  \\
        \tablemiddleline{}%
        Networks        & Framework     & Determines the sequence
                                          of transitions              \\
                        &               & or actions that
                                          constitutes a valid input.  \\
        Transitions     & Actions       & Assemble data and
                                          impose conditions the       \\
                        &               & input must meet to be
                                          accepted as valid.          \\
        Abbreviations   & Rules         & Responsible for
                                          recognising inputs.         \\
        \tablebottomline{}%
    \end{tabular}
\end{table}

@@

The framework manages the parsing process and provides support functions
for the actions, freeing the programmers writing actions to concentrate on
writing productive code.  It links actions and rules, allowing either to be
improved independently of the other, and allows new rules to be written
without needing changes to the source code of a parser.  The framework is
the core of the architecture and is deliberately quite simple: the rules
deal with the variation in inputs, and the actions deal with the
intricacies and complications encountered during parsing.  Finding the rule
that recognises the input is a very simple process, as shown by the
pseudo-code in \figureref{Pseudo-code describing how inputs are
recognised}.  The framework tries each rule until it finds one that
recognises the input, then it invokes the action specified by the rule.
The framework issues a warning if the input is not recognised by any of the
rules.

\begin{figure}[thbp]
    \caption{Pseudo-code showing how the framework finds the rule that
    recognises the input}
    \empty{}\label{Pseudo-code describing how inputs are recognised}
\begin{verbatim}
INPUT:
for each input {
    for each rule defined by the user {
        if this rule recognises the input {
            invoke the action specified by the rule
            next INPUT
        }
    }
    warn the user that the input was not recognised
}
\end{verbatim}
\end{figure}

@@

    \item [Register actions]  Each action needs to be registered with the
        framework so that the framework knows about it: the list of
        registered actions will be used when validating rules.

    \item [Load and validate rules]  The framework loads the rules from
        wherever they are stored: a simple file, a database, or possibly
        even a web server or other network service --- though that would
        have serious security implications.  It validates each rule to
        catch problems as early in the parsing process as possible; the
        checks will be implementation-specific to some extent, but will
        generally include the following:

@@

        Some optimisation steps may also be performed during loading of
        rules, as described in \sectionref{parser efficiency}.

    \item [Convert physical inputs to logical inputs] Each rule recognises
        a single input at a time: there is no facility for rules to consume
        more input or push unused input back onto the input stream,
        although actions may use cascaded parsing (explained in
        \sectionref{actions in architecture}) to push input back onto the
        input stream.  A physical input (e.g.\ a single line read from the
        input stream) may contain multiple or partial logical inputs, and
        the framework must pre-process these to provide a logical input for
        the rules to recognise.  This simplifies the rules and actions
        considerably, at the cost of added complexity in the framework;
        during the design phase it was decided that it was easier to deal
        with the problem of parsing multiple or partial inputs once, rather
        than dealing with it in every rule and action.  This is trivial for
        Postfix log files because they have a one-to-one mapping between
        physical and logical inputs; mapping between physical and logical
        inputs may be more difficult for other types of input.  Some input
        types may require pre-processing equivalent to parsing the physical
        inputs; in such cases the framework should take the approach
        adopted by many other parsers: combine the physical inputs into one
        complete input, use the rules to recognise the start of the input,
        discard the recognised portion if successful, and repeat until the
        input has been exhausted.

    \item [Classify the input]  The pseudo-code in \figureref{Pseudo-code
        describing how inputs are recognised} shows how rules are
        successively tried until one is found that recognises the input.
        That pseudo-code is very simple: there may be efficiency concerns
        (\sectionref{parser efficiency}), rule conditions
        (\sectionref{attaching conditions to rules}), or rule priorities
        (\sectionref{rules in architecture}) that complicate the process.

    \item [Invoke actions]  Once a rule has been found that recognises the
        input, the specified action will be invoked.  The framework
        marshals the data extracted by the rule, invokes the action, and
        pushes the modified input onto the input stream if the action uses
        cascaded parsing (see \sectionref{actions in architecture}).

    \item [Shared storage]  Parsers commonly need to save some state
        information about the input being parsed, e.g.\ a compiler tracking
        which variables are in lexical scope as it moves from one lexical
        block to another.  The framework provides shared storage to deal
        with this and any other storage needs the actions may have.
        Actions may need to exchange data to correctly parse the input,
        e.g.\ setting or clearing flags, maintaining a list of previously
        used identifiers, or ensuring at a higher level that the input
        being parsed meets certain requirements.

    \item [Save and load state]  The architecture can save the contents of
        the shared storage it provides for actions, and reload it later so
        that information is not lost between parsing runs.  \parsername{}
        does this because mails may take some time to deliver
        and thus have their log lines split between log files; a compiler
        might store data structures it builds as it parses different files.

    \item [Specialised support functions]  Actions may need support or
        utility functions; the framework may be a good location for support
        functions, but if another way exists to make those functions
        available to all actions it may be preferable to use that way
        instead, maintaining a clear separation of concerns.

@@

Each action is a separate procedure written to process a particular
category of input, e.g.\ rejection of a delivery attempt.  One input
category may have many input variants; in general each action will handle
one input category, with each rule recognising one input variant.  It is
anticipated that parsers based on this architecture will have many actions,
and each action may be invoked by many rules, with the aim of having simple
rules and tightly focused actions.  An action may need to process different
input variants in slightly different ways, but large variation in the
processing performed by an action indicates the need for a new action and a
new category of input; if an action becomes overly complicated it starts to
turn into a monolithic parser, with too much logic contained in a single
procedure.

The ability to easily add special purpose actions to deal with difficulties
and new requirements that are discovered during parser development is one
of the strengths of this architecture.  When a new requirement arises an
independent action can be written to satisfy it, instead of having to
modify a single monolithic function that processes every input, with all
the attendant risks of adversely affecting the existing parsing.  Sometimes
the new action will require the cooperation of other actions, e.g.\ to set
or check a flag, so actions are not always self-contained, but there will
still be a far lower degree of coupling and interdependency than in a
monolithic parser.

During development of \parsername{} it became apparent that in addition to
the obvious variety in log lines there were many complications to overcome.
Some were the result of deficiencies in Postfix's logging, and some of
those deficiencies were rectified by later versions of Postfix, e.g.\
identifying bounce notifications (\sectionref{identifying bounce
notifications}); others were due to the vagaries of process scheduling,
client behaviour, and administrative actions.  All were successfully
accommodated in \parsername{}: adding new actions was enough to overcome
several of the complications; others required modifications to a single
existing action to work around a difficulty; the remainder were resolved by
adapting existing actions to cooperate and exchange extra data (via the
framework), changing their behaviour as appropriate based on that extra
data.  Every architecture should aim to make the easy things easy and the
hard things possible; the successful implementation of \parsername{}
demonstrates that this architecture achieves that aim.

@@

This section is quite brief, because the architecture deliberately imposes
as few restrictions, conditions, and conventions as possible on actions, to
allow maximum flexibility for parsers based on this architecture.

@@

Rules are responsible for recognising inputs: each rule should recognise
one and only one input variant; an input category with multiple input
variants should have multiple rules, one for each variant.  Rules will
typically use a regex when recognising inputs, but other approaches may
prove useful for some applications, e.g.\ comparing fixed strings to the
input, or checking the length of the input; for the remainder of this
thesis it will be assumed that a regex is used.  Each rule must specify the
regex to recognise inputs and the action to invoke when recognition is
successful, but implementations are free to add any other attributes they
require; \sectionref{rules table} describes the attributes used in
\parsername{}, and some generally useful attributes will be discussed later
in this section.

Using the rules is simple: the first rule to recognise the input determines
the action that will be invoked; there is no backtracking to try alternate
rules, and no attempt is made to pick a \textit{best\/} rule.
\sectionref{attaching conditions to rules} contains an example which
requires that the rules are used in a specific order to correctly parse the
input, so a mechanism is needed to allow the author of the rules to specify
that ordering.  Each rule can have a priority attribute: when recognising
inputs, the framework should try the rules in the order specified by their
priority attributes, giving the ruleset author fine-grained control over
the order that rules are used in.  The priority attribute may be
implemented as a number, or as a range of values, e.g.\ low, medium, and
high, or in a different fashion entirely if it suits the implementation.
Rule ordering for efficiency is a separate topic that is covered in
\sectionref{rule ordering for efficiency}; overlapping rules are discussed
in \sectionref{overlapping rules in architecture}.

@@

Decoupling the rules from the actions and framework enables other rule
management approaches to be used, e.g.\ instead of manually editing
existing rules or adding new rules, machine learning techniques could be
used to automate the process.  If this approach was taken, the choice of
machine learning technique would be constrained by the size of typical data
sets (see \sectionref{parser efficiency}).  Techniques requiring the full
data set when training would be impractical; Instance Based
Learning~\cite{instance-based-learning} techniques that automatically
determine which inputs from the training set are valuable and which inputs
can be discarded might reduce the data required to a manageable size.  A
parser could also dynamically create new rules in response to certain
inputs, e.g.\ parsing a subroutine declaration could cause a rule to be
created that parses calls to that subroutine, checking that the arguments
used agree with the subroutine's signature.  These avenues of research and
development have not been pursued by the author, but the architecture
allows them to easily be undertaken independently.

@@

Rules can have conditions attached that will be evaluated by the framework
before attempting to use a rule to recognise an input: if the condition is
true the rule will be used, if not the rule will be skipped.  Conditions
can be as simple or complex as the parser requires, though naturally as the
complexity rises so too does the difficulty in understanding how different
rules and actions interact.  The framework has to evaluate each condition,
so as the complexity of conditions increases so will the complexity of the
code required to evaluate them.  Beyond a certain level of complexity,
conditions should probably be written in a proper programming language,
e.g.\ taking advantage of dynamic languages' support for evaluating code at
run-time, or embedding a language like Lua into the parser.  If an
implementation is going to use conditions so complex that they will require
a Turing-complete programming language, the design may need to be
revisited, including the decision to use this architecture --- there may be
other architectures more suitable.

Conditions that only examine the input will be the easiest conditions to
understand, because they can be understood in isolation; they do not depend
on variables set by actions or other rules.  Conditions that examine the
input can be complex if required, but simple conditions can be quite useful
too, e.g.\ every Postfix log line contains the name of the Postfix
component that produced it, so every rule used in \parsername{} has a
condition specifying the component whose log lines it recognises, reducing
the number of rules that will be used when recognising a log line (see
\sectionref{rule conditions in implementation} for details) and increasing
the chance that the log line will be correctly recognised.

@@

An example of how rule conditions can be used is parsing C-style comments,
which start with \texttt{/*} and end with \texttt{*/}; the start and end
tokens can be on one line, or may have many lines between them.
\Tableref{Rules to parse C-style comments} shows the regexes, conditions,
and state changes of the four rules required to parse C-style comments.
These are simplified rules, e.g.\ rules one and two will incorrectly
recognise the comment start token if it is within a quoted string.  Rules
one and two will be used when the parser is parsing code, not comments:
rule one recognises a comment that is contained within one line and leaves
the parser's state unchanged; rule two recognises the start of a comment
and changes the parser's state to parsing comments instead of parsing code.
Rules three and four will be used when the parser is parsing a comment:
rule three recognises the end of a comment and switches the parser's state
back to parsing code; rule four recognises a comment line without an end
token and keeps the parser's state unchanged.  It is important that the
rules are applied in the order listed in \tableref{Rules to parse C-style
comments} because rule two overlaps with rule one, and rule four overlaps
with rule three; \sectionref{rules in architecture} has explained how this
is achieved.  \sectionref{overlapping rules in architecture} will discuss
the benefits and difficulties of using overlapping rules.

\begin{table}[thbp]
    \caption{Rules to parse C-style comments}
    \empty{}\label{Rules to parse C-style comments}
    \begin{tabular}{llll}
        \tabletopline{}%
        No.   & Regex             & Condition                 & Variable Changes          \\
        \tablemiddleline{}%
        1     & \verb!/\*.*?\*/!  & state == parsing code     &                           \\
        2     & \verb!/\*.*!      & state == parsing code     & state = parsing comment   \\
        3     & \verb!.*\*/!      & state == parsing comment  & state = parsing code      \\
        4     & \verb!.*!         & state == parsing comment  &                           \\
        \tablebottomline{}%
    \end{tabular}

    Note that ``state'' is merely a descriptive name for the variable; the
    variable can be called anything at all.

@@

When adding new rules, the rule author must be aware that the new rule may
overlap with one or more existing rules, i.e.\ some inputs could be parsed
by more than one rule.  Unintentionally overlapping rules lead to
inconsistent parsing and data extraction because the first rule to
recognise the input wins, and the order in which rules are used might
change between parser invocations.  Overlapping rules are frequently a
requirement, allowing a more specific rule to recognise some inputs and a
more general rule to recognise the remainder, e.g.\ separating
\acronym{SMTP} delivery to specific sites from \acronym{SMTP} delivery to
the rest of the world.  Using overlapping rules simplifies both the general
rule and the more specific rule.  Overlapping rules should have a priority
attribute to specify their relative ordering; negative priorities may be
useful for catchall rules.  The architecture does not try to detect
overlapping rules: that responsibility is left to the author of the rules.

Overlapping rules can be detected by visual inspection, or a program could
be written to analyse the regexes in a ruleset.  Traditional regexes are
equivalent in computational power to \acronym{FA} and can be converted to
\acronym{FA}, so regex overlap can be detected by finding a non-empty
intersection of two \acronym{FA}\@.  Perl 5.10 regexes~\cite{perlre} are
more powerful than traditional regexes: they can match correctly balanced
brackets nested to an arbitrary depth, e.g.\
\verb!/^[^<>]*(<(?:(?>[^<>]+)|(?1))*>)[^<>]*$/!  matches
\verb!z<123<pq<>rs>j<r>ml>s!.  Matching balanced brackets requires the
regex engine to maintain state on a stack, so Perl 5.10 regexes are
equivalent in computational power to \acronym{PDA}; detecting overlap may
require calculating the intersection of two \acronym{PDA} instead of two
\acronym{FA}.  \acronym{PDA} intersection is not closed, i.e.\ the result
cannot always be implemented using a \acronym{PDA}, so intersection may be
intractable sometimes, e.g.:
$a^{*}b^{n}c^{n}~\cap~a^{n}b^{n}c^{*}~\rightarrow~a^{n}b^{n}c^{n}$.
Detecting overlap amongst $n$ regexes requires calculating
$\frac{n\left(n-1\right)}{2}$ intersections, resulting in
$O\left(n^{2}x\right)$ complexity, where $x$ is the cost of calculating
\acronym{FA} or \acronym{PDA} intersection.  This is certainly not a task
to be performed every time a parser runs: naive detection of overlap
amongst \parsernames{} \numberOFrules{} rules would require calculating
\numberOFruleINTERSECTIONS{} intersections.

@@

In the worst case, $|o|$ above will be equal to $n^{2}$.  For this approach
to be more efficient than the naive approach, $y$ must be significantly
lower than $x$.  If $y$ is higher than $x$ then the checks for overlap
should be performed in the opposite order: check for regex overlap first,
then check for condition overlap only between pairs of rules with
overlapping regexes.

@@

It is possible to write pathological regexes, which fall into two main
categories: regexes that match inputs they should not, and regexes that
consume excessive amounts of CPU time during matching.  Defining a regex
that matches inputs it should not is trivial: \verb!/^/! matches the start
of every input.  This regex would be found by a tool that detects
overlapping rules, and would easily be noticed by visual inspection, but
more complex regexes would be harder to find.  Regexes that match inputs
more than they should are a problem not because of excessive resource
usage, but because they may prevent the correct rule from recognising the
input.  If an adaptive ordering system is used to prioritise rules that
frequently recognise inputs (see \sectionref{rule ordering for
efficiency}), then a rule with a regex that matches inputs it should not
may be promoted up through the list, displacing an increasing number of
correct rules as it rises.
